Unsupervised embedding of trajectories captures the latent structure of mobility

### 2021-02-18 paper/rebuttal/rebuttal.tex

In other words, \emph{word2vec with the negative sampling is a gravity model}, with the mass given by the location's frequency $P(i)$, and the distance measured by their dot similarities.
While the gravity model generates the flow from the given mass and locations, {\it word2vec} finds the mass and locations that best explain the given trajectories.
Therefore, the embedding distance is inherently tied to the flow, and hence, has greater predictive power than the geographic distance.

A central assumption in our mathematical analysis is that the embedding vectors are optimum in terms of the {\it word2vec}'s objective function.
However, this premise is not practical because the optimization algorithm (\eg stochastic gradient descent algorithm) often finds sub-optimal embeddings due to its stochasticity and being trapped in a local optimum.
This has two consequences.
First, the in- and out-vectors can be different.
Second, dot similarity should in principle predict the mobility flow better than cosine similarity because dot similarity includes cosine similarity as its special case.
However, our simulations show that it is the cosine similarity that better predicted the mobility flow~(Tables~\ref{supp:table:r2_table},~\ref{supp:table:rmse_exp_table},~\ref{supp:table:rmse_power_table}).
It has been widely observed that cosine similarity (dot product with $l2$ normalization) is generally better than dot similarity and dot similarity with various normalization methods in capturing a similarity between two words ~\autocite{levy2015improving}.
Nevertheless, among the ten measures tested in our simulations, dot similarity is the second-best predictor of mobility flow following cosine similarity (Tables~\ref{supp:table:r2_table},~\ref{supp:table:rmse_exp_table},~\ref{supp:table:rmse_power_table}).
%What makes cosine similarity superior to dot similarity warrants the future work.

\subsection{Changes to the manuscript}
To reflect these breakthroughs, we have made the following changes to the manuscript:
\begin{itemize}
	\itemsep0em
	\item Following results demonstrating the performance of the embedding distance compared to geographic distance, we add the section \textbf{word2vec and the gravity model}, which summarizes the link between the two models.
	\item We also added the section \textbf{An implicit bias in the negative sampling} to the methods section of our manuscript.
	\item Since the formulation of the gravity used in the above derivations assumes $\gamma = 1$, we use this value of $gamma$ for all reported results. No result substantively changes with this change.
	\item We added a sentence to the paper speculating on why cosine similarity works best in the main text.
	\item We also added a sentence noting the in-vector and out-vector issue, and speculating on reasons/consequences.
\end{itemize}

@@

\subsection{Why do neural embeddings outperform direct matrix factorization?}% why sgd is better?
One question raised by these comparisons is, why do neural-embedding approaches, which rely on stochastic gradient descent, outperform direct matrix factorization, especially seeing as \textit{word2vec} implicitly approximates factorization?
An issue in the matrix factorization approach is that it is extremely sensitive to small flows between affiliations. The matrix factorization approach embeds affiliations $i$ and $j$ such that their dot similarity is as close as to $\log (T_{ij} / P(i) P(j))$. If the flow $T_{ij}$ is considerably small or zero, the dot similarity goes $-\infty$, pushing $i$ and $j$ too far from the other affiliations in the embedding space~\autocite{levy2014neural, Qui2018}. This is particularly problematic when the window size is small because most affiliation pairs would have no flow, which is indeed the case in our experiments. To circumvent this problem, the previous studies~\autocite{levy2014neural, Qui2018} add a constant flow between the affiliation pairs with no flow. However, this changes the data to which we fit the model, and small flows still have a considerably strong impact on the embedding.

Also, it is well known that SVD by itself is very vulnerable to outliers \autocite{ xu2012robust, huber1981robust, xu1995robust, chandrasekaran2011rank, candes2011robust, }. Compared to SVD, a stochastic gradient descent algorithm (SGD) which is employed in SGNS \textit{word2vec} benefits from the noise in the model itself and can enhance generalization and effectiveness as well \autocite{ma2018power, smith2020generalization, zhang2019algorithmic,}.

\subsection{Changes to the manuscript}
To reflect these additional experiments, we have made the following changes to our manuscript:
\begin{itemize}
	\item Added Fig~\ref{fig:supp:distancemetrics} and Tables~\ref{supp:table:r2_table},~\ref{supp:table:rmse_exp_table},~\ref{supp:table:rmse_power_table} to the manuscript.
	\item We have added sentence speculating on SGD performance differences into supporting materials
\end{itemize}

@@

\response{Thank you for pointing this out.
	We carefully read out the Barthélemy, Physics Reports 499, 1, 2011, and referenced it in our manuscript.
	Also, we report the performance of the flux estimation and prediction using different deterrence functions (see Section 2, above), finding that while performance differs, the embedding distance outperforms geographic distance no matter the function used.
	We add the following line to the methods noting this point,

	\medskip
	\say{The decay function $f(r_{ij})$ defines the effect of distance, and so different decay functions can model fundamentally different mechanism [...] such as the cost functions for a given distance and the spatial granularity of the observation.
		For geographic distance, we define $f(r_{ij})$ as the standard power-law function, and for the embedding distance, we use the exponential function, selected as the best performing for each case though the embedding distance outperforms geographic distance regardless of the type of gravity model used (see Supporting Information).}

	We find the article that geographical distance ﬁts well with the power-law decay function because power-based gravity model can be more suitable for analyzing large, complex, and scale-free spatial systems theoretically and the exp`onential-based gravity model indicates simplicty rather than spatial complexity~\autocite{chen2015distance}. Meanwhile, embedding distance ﬁts well with the exponential decay function, which we demonstrate to correspond to an underlying mathematical relationship between \textit{word2vec} and the gravity model, a connection we recently discovered and included as a major finding in this revision (see our response to Referee 1).

### 2021-02-18 paper/current/main.tex
re-edit the last paragraph of math

A central assumption in our mathematical analysis is that the embedding vectors are optimum in terms of the {\it word2vec}'s objective function.
However, this premise is not practical because the optimization algorithm (\eg stochastic gradient descent algorithm) often finds sub-optimal embeddings due to its stochasticity and being trapped in a local optimum.
This has two consequences.
First, the in- and out-vectors can be different.
Second, dot similarity should in principle predict the mobility flow better than cosine similarity because dot similarity includes cosine similarity as its special case.
However, our simulations show that it is the cosine similarity that better predicted the mobility flow~(Tables~\ref{supp:table:r2_table},~\ref{supp:table:rmse_exp_table},~\ref{supp:table:rmse_power_table}).
It has been widely observed that cosine similarity (dot product with $l2$ normalization) is generally better than dot similarity and dot similarity with various normalization methods in capturing a similarity between two words ~\autocite{levy2015improving}.
Nevertheless, among the ten measures tested in our simulations, dot similarity is the second-best predictor of mobility flow following cosine similarity.

>>>>

A central assumption in our mathematical analysis is that the embedding vectors are optimum in terms of the {\it word2vec}'s objective function.
However, this premise is not practical because the optimization algorithm (\eg stochastic gradient descent algorithm) often finds sub-optimal embeddings.
This has two consequences.
First, the in- and out-vectors can be different.
Second, dot similarity calculated by {\it word2vec} is the distance metric tied to the gravity model, and hence, should capture the mobility flow well.
However, in our numerical simulations, cosine similarity calculated by {\it word2vec} predicts better than the dot similarity~(Tables~\ref{supp:table:r2_table},~\ref{supp:table:rmse_exp_table},~\ref{supp:table:rmse_power_table}).
Cosine similarity is a special case of dot similarity, and thus, should be no better than dot similarity if the embedding vectors are optimal.
In practice, cosine similarity often outperform dot similarity in many applications of word embedding such as word analogy test~\autocite{levy2015improving}.
Nevertheless, among the ten measures tested in our simulations, dot similarity is the second-best predictor of mobility flow, can be even better than cosine similarity.
